Covariate and analysis of MSBB reprocessed counts Regressing out Diagnosis as well for TWAS

### 2019-11-09 code/MSBB_RegressDiagnosis_4_TWAS_hg38.Rmd
hg38 Update

syn8691099.1
syn10156693.3
syn8698270.1
syn8449369.2
syn6100548.7
syn6101474.4

@@

### Data download
Obtain count matrix and metadata from synapse
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn8691099';
ALL_USED_IDs <- COUNT_ID
COUNT <- syn_temp$get(COUNT_ID, version = 1)$path %>%
  read.table(header=T, sep='\t', check.names = F, row.names = 1)
# Convert rownames of counts from tracking id to ensemble gene id
tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
  dplyr::mutate(ID = Gene.ID) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(tmp) = tmp$Gene.ID
rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']
# Get sample ids
SampleID = data.frame(SampleID = colnames(COUNT),
                      ID = colnames(COUNT)) %>% 
  tidyr::separate(ID, c('A','B','ID'), sep = '_') %>%
  dplyr::select(SampleID, ID) %>%
  dplyr::mutate(ID = as.numeric(ID))
# Get technical metadata
METADATA_TECH_ID <- 'syn6100548'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
METADATA_TECH <- syn_temp$get(METADATA_TECH_ID, version = 7)$path %>%
  read.table(sep=',',header=T, row.names=1) %>%
  group_by(sampleIdentifier) %>%
  top_n(1, -rRNA.rate) %>%
  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN) %>%
  unique %>%
  ddply(.(barcode), .fun = function(x){
    x = dplyr::mutate(x, batch = paste(batch, collapse = ''), RIN = sum(RIN, na.rm = T))  
  }) 
# Get 90+ age
METADATA_AGE_ID = 'syn10156693'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_AGE_ID
METADATA_AGE <- syn_temp$get(METADATA_AGE_ID, version = 3)$path %>%
  read.table(sep='\t', header=T)
# Get clinical metadata
METADATA_CLINICAL_ID <- 'syn6101474'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- syn_temp$get(METADATA_CLINICAL_ID, version = 4)$path %>%
  read.csv(header = T)
# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8698270';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD <- syn_temp$get(METADATA.PICARD_ID, version = 1)$path %>%
  read.table(sep='\t', header=T, stringsAsFactors=F) %>% 
  dplyr::rename(sampleIdentifier = sample)
METADATA <- dplyr::full_join(METADATA_TECH, METADATA_CLINICAL) %>%
  dplyr::inner_join(METADATA.PICARD) %>%
  dplyr::select(-AOD) %>%
  dplyr::left_join(METADATA_AGE)
colnames(METADATA) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA))
colnames(METADATA) = gsub('RnaSeqMetrics__','',colnames(METADATA))
# Get gene specific parameters from synapse
GENE.PARAM = syn_temp$get('syn8449369', version = 2)$path %>% 
    read.table(sep='\t', header=T, stringsAsFactors=F)
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')
GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id
```

@@

### Filter genes
* Remove genes that have less than 1 cpm counts in at least 50% of samples per Tissue x Diagnosis
* Remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Tissue.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
  }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
# Check gene biotype
## Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
## Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
                       filters = "ensembl_gene_id", 
                       values = PROCESSED_COUNTS$filteredExprMatrix$genes,
                       mart = mart)
summary(factor(Ensemble2HGNC$gene_biotype)) %>%
  rownameToFirstColumn('Biotype') %>%
  dplyr::rename(fraction = DF) %>%
  dplyr::mutate(fraction = fraction/dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]) %>%
  dplyr::filter(fraction >= 0.01) %>%
  kable
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn20798011';
activityName = 'Covariate and Diagnosis Regression';
activityDescription = 'Covariate analysis and Regrsison of AD Diagnosis of aligned effective counts with GRCh37 with CQN normalisation (IFG, STG, FP, PHG)';
thisFileName <- 'MSBB_RegressDiagnosis_4_TWAS.Rmd'
# Github link
thisRepo <- githubr::getRepo(repository = "jgockley62/TWAS", ref="branch", refName='master')
thisFile <- githubr::getPermlink(repository = thisRepo, repositoryPath=paste0('code/',thisFileName))
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
activityName = 'Covariate and Diagnosis Regression';
activityDescription = 'Covariate analysis and Regrsison of AD Diagnosis of aligned effective counts with GRCh37 with CQN normalisation (IFG, STG, FP, PHG)';

@@

#Set Used SynIDs For Provenance
Syns_Used <- c("syn8691099", "syn10156693", "syn8698270", 
               "syn8449369", "syn6100548", "syn6101474"
)
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'RNAseq',
  tissueTypeAbrv	= c('IFG', 'STG', 'FP', 'PHG'), 
  study = 'MSBB', 
  organism = 'HomoSapiens',
  consortium	= 'AMPAD',
  normalizationStatus	= TRUE,
  normalizationType	= 'CQN',
  rnaquantification = 'RSEM',
  genomeAssemblyID = 'GRCh37'
)

@@

createAndKnitToFolderEntityClient(file = "MSBB_RegressDiagnosis_4_TWAS.Rmd",
                                          parentId ="syn20798011",
                                          folderName = 'MSBB')
```

>>>>

| *Synapse ID* | *File Name* |
|  -----------------------------------------  |   ---------                      |
| Old Count Matrix ------syn8691099.1 | |
| syn20963389 | |
| syn20963390 | |
| syn10156693.3 | | 
| syn8698270.1 | | 
| Old Length GC Data-----syn8449369.2 | | 
| syn21156161 | |
| syn6100548.7 | | 
| syn6101474.4 | |

@@

### Data download
Obtain count matrix and metadata from synapse
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn20963390';
ALL_USED_IDs <- COUNT_ID
COUNT <- syn_temp$get(COUNT_ID, version = 1)$path %>%
  read.table(header=T, sep='\t', check.names = F, row.names = 1)

Provenance <- 'syn20963389'
Prov <- syn_temp$get(Provenance, version = 1)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = Provenance
Translator <- read.csv(Prov$path, header=T, row.names = 1)
colnames(COUNT) <- as.character( Translator[ colnames(COUNT), ][,2] )

COUNT <- COUNT[ row.names(COUNT)[ (row.names(COUNT) %in% c('N_unmapped', 'N_multimapping', 'N_noFeature', 'N_ambiguous'))==F] , ]

#Need to filter low counts here to remove the PAR_Y Regions
COUNT <- COUNT[ rowSums(COUNT[,] > 0) / (ncol(COUNT)) >= 0.05, ]

# Convert rownames of counts from tracking id to ensemble gene id
tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
  dplyr::mutate(ID = Gene.ID) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(tmp) = tmp$Gene.ID
rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']
# Get sample ids
SampleID = data.frame(SampleID = colnames(COUNT),
                      ID = colnames(COUNT)) %>% 
  tidyr::separate(ID, c('A','B','ID'), sep = '_') %>%
  dplyr::select(SampleID, ID) %>%
  dplyr::mutate(ID = as.numeric(ID))
# Get technical metadata
METADATA_TECH_ID <- 'syn6100548'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
METADATA_TECH <- syn_temp$get(METADATA_TECH_ID, version = 7)$path %>%
  read.table(sep=',',header=T, row.names=1) %>%
  group_by(sampleIdentifier) %>%
  top_n(1, -rRNA.rate) %>%
  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN) %>%
  unique %>%
  ddply(.(barcode), .fun = function(x){
    x = dplyr::mutate(x, batch = paste(batch, collapse = ''), RIN = sum(RIN, na.rm = T))  
  }) 
# Get 90+ age
METADATA_AGE_ID = 'syn10156693'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_AGE_ID
METADATA_AGE <- syn_temp$get(METADATA_AGE_ID, version = 3)$path %>%
  read.table(sep='\t', header=T)
# Get clinical metadata
METADATA_CLINICAL_ID <- 'syn6101474'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- syn_temp$get(METADATA_CLINICAL_ID, version = 4)$path %>%
  read.csv(header = T)
# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8698270';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD <- syn_temp$get(METADATA.PICARD_ID, version = 1)$path %>%
  read.table(sep='\t', header=T, stringsAsFactors=F) %>% 
  dplyr::rename(sampleIdentifier = sample)
METADATA <- dplyr::full_join(METADATA_TECH, METADATA_CLINICAL) %>%
  dplyr::inner_join(METADATA.PICARD) %>%
  dplyr::select(-AOD) %>%
  dplyr::left_join(METADATA_AGE)
colnames(METADATA) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA))
colnames(METADATA) = gsub('RnaSeqMetrics__','',colnames(METADATA))
# Get gene specific parameters from synapse
GENE.PARAM = syn_temp$get('syn21156161', version = 3)$path %>% 
    read.table(sep='\t', header=T, stringsAsFactors=F)
ALL_USED_IDs = c(ALL_USED_IDs, 'syn21156161')
GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id
```

@@

### Filter genes
* Remove genes that have less than 1 cpm counts in at least 50% of samples per Tissue x Diagnosis
* Remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Tissue.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
  }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
# Check gene biotype
## Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2019.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
## Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
                       filters = "ensembl_gene_id", 
                       values = PROCESSED_COUNTS$filteredExprMatrix$genes,
                       mart = mart)
summary(factor(Ensemble2HGNC$gene_biotype)) %>%
  rownameToFirstColumn('Biotype') %>%
  dplyr::rename(fraction = DF) %>%
  dplyr::mutate(fraction = fraction/dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]) %>%
  dplyr::filter(fraction >= 0.01) %>%
  kable
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn21136908';
activityName = 'Covariate and Diagnosis Regression';
activityDescription = 'Covariate analysis and Regrsison of AD Diagnosis of aligned effective counts with GRCh38 with CQN normalisation (IFG, STG, FP, PHG)';
thisFileName <- 'MSBB_RegressDiagnosis_4_TWAS_hg38.Rmd'
# Github link
thisRepo <- githubr::getRepo(repository = "jgockley62/TWAS", ref="branch", refName='master')
thisFile <- githubr::getPermlink(repository = thisRepo, repositoryPath=paste0('code/',thisFileName))
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
activityName = 'Covariate and Diagnosis Regression';
activityDescription = 'Covariate analysis and Regrsison of AD Diagnosis of aligned effective counts with GRCh38 with CQN normalisation (IFG, STG, FP, PHG)';

@@

#Set Used SynIDs For Provenance
Syns_Used <- c("syn20963390", "syn10156693", "syn8698270", 
               "syn21156161", "syn6100548", "syn6101474", "syn20963389"
)
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'RNAseq',
  tissueTypeAbrv	= c('IFG', 'STG', 'FP', 'PHG'), 
  study = 'MSBB', 
  organism = 'HomoSapiens',
  consortium	= 'AMPAD',
  normalizationStatus	= TRUE,
  normalizationType	= 'CQN',
  rnaquantification = 'RSEM',
  genomeAssemblyID = 'GRCh38'
)

@@

createAndKnitToFolderEntityClient(file = "MSBB_RegressDiagnosis_4_TWAS_hg38.Rmd",
                                          parentId ="syn21136908",
                                          folderName = 'MSBB')
```
